\LARGE \bf Mean Estimation from One-Bit Measurements

### 2019-12-13 adaptive-estimation.tex
Giving the one step estimator proof

As an immediate corollary to Theorem~\ref{thm:sgd}, we obtain the following
asymptotic optimality results of the averaged stochastic gradient
sequence.
\begin{corollary}
  Let the conditions of Theorem~\ref{thm:adpative_lower_bound} hold
  and $\theta_n$ be defined by the iteration~\eqref{eq:sgd_alg}.
  Let $\{P_\theta\}_{\theta \in \R}$ be the family of distributions
  with densities $f(\cdot - \theta)$.
  \begin{enumerate}[(i)]
  \item Define the shorthand $P_n = P_{\theta + h_n/\sqrt{n}}^n$.
    If the stepsizes satisfy condition~\eqref{eqn:lazy-gamma}, then
    \begin{equation*}
      \sqrt{n}(\bar{\theta}_n - \theta - h_n / \sqrt{n})
      \mathop{\cd}_{P_n} \normal\left(0, \frac{1}{\eta(0)}\right).
    \end{equation*}
  \item If in addition the stepsizes satisfy
    condition~\eqref{eqn:stringent-gamma}, then for any
    prior $\pi$ on $\R$, we have
    \begin{equation*}
      \int \E\left[(\bar{\theta}_n - \theta)^2\right] d\pi(\theta)
      = \frac{1}{4 n f(0)^2} + o(n^{-1}).
    \end{equation*}
  \end{enumerate}
\end{corollary}
\noindent

Theorem~\ref{thm:sgd} implies that the estimator ${\theta}_n$ defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg} attains the maximal ARE, as established by Theorem~\ref{thm:adpative_lower_bound}. %Furthermore, it is locally minimax in the sense that it attains the minimal MSE for all local alternatives of $\theta$. 
%
The update step \eqref{eq:sgd_alg} can be seen as a gradient descent step for the function $x\to |x|$ at the point $x=X_n - \theta_{n-1}$. Consequently, the procedure above is known as averaged stochastic gradient descent for minimizing $x \to |x|$ given the data $X_1,\ldots,X_n$. The minimal value of this optimization is the sample median, and Theorem~\ref{thm:sgd} provides conditions for the sequence of gradient steps so that the algorithm converges to this minimum. In particular, parts (i) and (ii) of Theorem~\ref{thm:sgd} provides conditions under which $\bar{\theta}$ is asymptotically normal and local asymptotically minimax, in the sense that it attains the local asymptotic minimax bound \cite{van2000asymptotic}. Part (iii) provides an additional condition under which $\bar{\theta}$ also converges in its second moment. 
 \par
%Note that $\theta_0$ is not explicitly defined in equation \eqref{eq:sgd_est}. A reasonable initialization for $\theta_0$ is $\theta_0 = \mathbb E [\theta]$, although Theorem~\ref{thm:sgd} implies that the asymptotic behavior of the estimator is indifferent to this initialization. Thus, the optimal efficiency is attained regardless of the prior distribution on $\theta$ or the radius of the parameter space $\Theta$.  \\
In the encoding and estimating procedure \eqref{eq:sgd_alg} and \eqref{eq:sgd_est}, each one-bit message $B_n$ depends on its private sample as well as the current gradient descent estimate $\theta_{n-1}$. In this sense, each encoder in this algorithm interacts with previous one by using the current estimate.
%
As we explain next, 
it is possible to obtain the optimal efficiency of $\eta(0)\sigma^2$ with only one round of interactions among the encoders.

@@

\subsection{Maximal Efficiency using One Round of Threshold Adaptation}
In Section~\ref{sec:preliminary} we considered an estimator that is based on binary messages of the form 
\begin{equation*}
B_i = \mathbf 1_{X_i > \theta_0},\qquad i=1,\ldots,n,
\end{equation*}
and deduced that it is asymptotically normal with variance $1/\eta(\theta-\theta_0)$.
%
We now show that a similar encoding leads to an asymptotically normal estimator attaining the lower variance bound $1/\eta(0)$, provided we allow to  update the threshold value $\theta_0$ based on previously observed bits at least once. 
%
In this procedure we separate the sample into two disjoint sets: $X_1,\ldots,X_{n_1}$ and $X_{n_1+1},\ldots,X_n$ for some $n_1 < n$.
We first use the estimator \eqref{eq:estimator_naive} to obtain an estimate ${\theta}_{n_1}$ based on $B_1,\ldots,B_{n_1}$, and then use ${\theta}_{n_1}$ as the new threshold value to obtain messages $B_{n_1+1}, \ldots, B_n$. Figure~\ref{fig:one_round} illustrates a diagram of this procedure. 
%
The specific encoding and estimation scheme, as well as its asymptotic performance, are given by the following theorem:
\begin{thm}
For $i=1,\ldots,n$ set
\begin{equation*}
B_i = \begin{cases}
 \mathbf 1_{X_i \geq \theta_0} & i = 1,\ldots,n_1, \\
 \mathbf 1_{X_i \geq {\theta}_{n_1} }& i={n_1+1,\ldots,n},
\end{cases}
\end{equation*}
where $n = n_1+ n_2$ and 
\begin{equation*}
{\theta}_{n_1} \triangleq \theta_0 + F^{-1}\left(
\frac{1}{n_1} \sum_{i=1}^{n_1} B_i 
 \right).
\end{equation*} 
Let 
\begin{equation*}
{\theta}_{n_2}  \triangleq {\theta}_{n_1} +  F^{-1} \left( \frac{1}{n_2} \sum_{i=n_1+1}^{n_2} B_i \right)
\end{equation*}
and assume that, as $n\to \infty$, $n_1(n) \rightarrow \infty$ and $n_2(n)/n \rightarrow 1$. Then:
\begin{align*}
 \sqrt{n} \left( {\theta}_{n_2} - \theta  \right) \overset{d}{\longrightarrow}  \Ncal\left( 0, 1/\eta(0) \right).
\end{align*}
\end{thm}
%
%
\begin{proof}
For $t\in \mathbb R$, set
\begin{equation*}
p_{n_2}(t) \triangleq \frac{1}{n_2} \sum_{i=n_1+1}^{n} \mathbf 1_{X_i \geq t}. 
\end{equation*}
Since $n_2(n)/n \to 1$, if follows from the Berry Esseen theorem that there exists $C$, independent of $n$ and $t$, such that 
\begin{align*}
\sup_{t \in \R} \left| \sqrt{n} \frac{ p_{n_2}(t) - F( \theta - t)} {F(\theta - t) F \left( t - \theta\right) }  - \Phi(t) \right|  \leq C/\sqrt{n}. 
\end{align*}
where $Z \sim \Ncal(0,1)$. It follows that
\begin{align}
\label{eq:two_step_proof}
 \sqrt{n} \frac{ p_{n_2}(Y_n) - F(0)} {F^2(0) }  \overset{d}{\rightarrow} \Ncal(0,1)
\end{align}
for any sequence of random variables $\{Y_n\}$ converging in probability to $\theta$. In particular, since
\begin{equation*}
p_{n_1} \triangleq \frac{1}{n_1} \sum_{i=1}^{n_1} B_i \overset{a.s.}{\rightarrow} F(\theta - \theta_0)
\end{equation*}
by the law of large numbers, \eqref{eq:two_step_proof} holds for $Y_n = \theta_{n_1}$. Finally, we use the delta method with $g(x) = F^{-1}(x)$ to conclude
\begin{align*}
& \sqrt{n}\left( \theta_{n_2} - \theta \right) = \sqrt{n} \left( \theta_{n_1} + F^{-1}(p_{n_2}(\theta_{n_1})) - \theta \right) \\
&= \sqrt{n} \left( g(p_{n_2}(\theta_{n_1})) - g \left(F(\theta-\theta_{n_1}) \right)  \right) \overset{d}{\to} \Ncal\left(0, 
F^2(0)/ f^2(0) \right).
\end{align*}
\end{proof}
%

>>>>

As an immediate corollary to Theorem~\ref{thm:sgd}, we obtain the following
asymptotic optimality results of the averaged stochastic gradient
sequence. Specifically, the average of the stochastic gradient
iterates~\eqref{eq:sgd_alg}
is locally asymptotically minimax, and they achieve the lower
bound of Theorem~\ref{thm:adpative_lower_bound}.

\begin{corollary}
  Let the conditions of Theorem~\ref{thm:adpative_lower_bound} hold
  and $\theta_n$ be defined by the iteration~\eqref{eq:sgd_alg}.
  Let $\{P_\theta\}_{\theta \in \R}$ be the family of distributions
  with densities $f(\cdot - \theta)$.
  \begin{enumerate}[(i)]
  \item Define the shorthand $P_n = P_{\theta + h_n/\sqrt{n}}^n$.
    If the stepsizes satisfy condition~\eqref{eqn:lazy-gamma}, then
    \begin{equation*}
      \sqrt{n}(\bar{\theta}_n - \theta - h_n / \sqrt{n})
      \mathop{\cd}_{P_n} \normal\left(0, \frac{1}{\eta(0)}\right).
    \end{equation*}
  \item If in addition the stepsizes satisfy
    condition~\eqref{eqn:stringent-gamma}, then they
    achieve the lower bound of Theorem~\ref{thm:adpative_lower_bound} for any
    prior $\pi$ on $\R$.
  \end{enumerate}
\end{corollary}

%% Theorem~\ref{thm:sgd} implies that the estimator ${\theta}_n$ defined by \eqref{eq:sgd_est} and \eqref{eq:sgd_alg} attains the maximal ARE, as established by Theorem~\ref{thm:adpative_lower_bound}. %Furthermore, it is locally minimax in the sense that it attains the minimal MSE for all local alternatives of $\theta$. 
%% %
%% The update step \eqref{eq:sgd_alg} can be seen as a gradient descent step for the function $x\to |x|$ at the point $x=X_n - \theta_{n-1}$. Consequently, the procedure above is known as averaged stochastic gradient descent for minimizing $x \to |x|$ given the data $X_1,\ldots,X_n$. The minimal value of this optimization is the sample median, and Theorem~\ref{thm:sgd} provides conditions for the sequence of gradient steps so that the algorithm converges to this minimum. In particular, parts (i) and (ii) of Theorem~\ref{thm:sgd} provides conditions under which $\bar{\theta}$ is asymptotically normal and local asymptotically minimax, in the sense that it attains the local asymptotic minimax bound \cite{van2000asymptotic}. Part (iii) provides an additional condition under which $\bar{\theta}$ also converges in its second moment.

@@

\subsection{Maximal Efficiency using One Round of Threshold Adaptation}

In the encoding and estimating procedure \eqref{eq:sgd_alg}, each one-bit
message $B_n$ depends on its private sample as well as the current gradient
descent estimate $\theta_{n-1}$. In this sense, each encoder in this
algorithm interacts with previous one by using the current estimate.  This
amount of adaptivity is unnecessary: as we now consider, a similar encoding
yields an asymptotically normal estimator attaining the lower variance bound
$1/\eta(0)$, provided we allow \emph{one} adaptive update to the threshold
value $\theta_0$ based on previously observed bits.
%
In this procedure we separate the sample into the disjoint sets
$X_1,\ldots,X_{n_1}$ and $X_{n_1+1},\ldots,X_n$ for some $n_1 < n$.  We
first use the estimator \eqref{eq:estimator_naive} to obtain an estimate
${\theta}_{n_1}$ based on $B_1,\ldots,B_{n_1}$, and then use
${\theta}_{n_1}$ as the new threshold value to obtain messages $B_{n_1+1},
\ldots, B_n$. Figure~\ref{fig:one_round} illustrates a diagram of this
procedure.

More formally, we consider the following estimation scheme. Given
$n_1 \in \{1, \ldots, n\}$,
set the individual bits
\begin{equation*}
  B_i =
  \begin{cases}
    \indic{X_i \geq \theta_0} & i = 1,\ldots,n_1, \\
    \indic{X_i \geq T_n}& i={n_1+1,\ldots,n},
  \end{cases}
\end{equation*}
where
\begin{equation*}
  T_n \defeq \theta_0 + F^{-1}\left(
  \frac{1}{n_1} \sum_{i=1}^{n_1} B_i 
  \right)
  ~~ \mbox{and} ~~
  \theta_n \defeq
  \theta_{n_1} + F^{-1} \left(\frac{1}{n - n_1}
  \sum_{i = n_1}^n B_i \right).
\end{equation*} 
Then
\begin{thm}
  Assume that as $n\to \infty$, $n_1(n) \rightarrow \infty$ and $n_1 / n \to
  0$. Then
  \begin{align*}
    \sqrt{n} \left( {\theta}_n - \theta  \right)
    \cd  \normal\left( 0, 1/\eta(0) \right).
  \end{align*}
\end{thm}
%
%
\begin{proof}
  We abuse notation and instead of assuming we receive $n$ observations, assume
  we receive the $n + n_1$ observations $X_{-n_1}, \ldots, X_{-1}$ and
  $X_1, \ldots, X_n$, defining $T_n = \theta_0 + F^{-1}(\frac{1}{n_1}
  \sum_{i = -n_1}^{-1} B_i)$ and
  $B_i = \indic{X_i \ge T_n}$ for $i \ge 1$.
  By continuity, we have $T_n \cas \theta$ as $n_1 \to \infty$.
  Let
  $E_n = \E[B_i \mid T_n] = P(X_i \ge T_n)$,
  so that $V_n \defeq \var(B_i \mid T_n) = E_n(1 - E_n)$. Define also the random
  variable
  \begin{equation*}
    Y_n \defeq \frac{1}{\sqrt{n}} \sum_{i = 1}^n \frac{B_i - E_i}{
      \sqrt{E_n (1 - E_n)}},
  \end{equation*}
  and let $F_n(\cdot \mid T_n)$ be its cumulative distribution function.
  Then because
  \begin{equation*}
    \E\left[|B_i - E_n|^3 \mid T_n\right] \le V_n = E_n(1 - E_n)
    ~~ \mbox{we have} ~~
    \E\left[\frac{|B_i - E_n|^3}{(E_n(1 - E_n))^{3/2}} \mid T_n\right]
    \le \frac{1}{\sqrt{V_n}}.
  \end{equation*}
  Then the Berry-Esseen theorem implies that there exists a constant
  $C \le 1$ such that
  \begin{equation*}
    \sup_t \left|F_n(t \mid T_n) - \Phi(t) \right|
    \le \frac{C}{\sqrt{V_n} \sqrt{n}} \wedge 2,
  \end{equation*}
  where $\Phi$ is the standard Gaussian CDF.
  As $V_n \cas \frac{1}{4}$ by definition of the median,
  we have
  that (with probability 1)
  \begin{equation*}
    \sup_t \left|F_n(t \mid T_n) - \Phi(t)\right| \le \frac{C}{\sqrt{n}}
    ~~ \mbox{eventually}.
  \end{equation*}
  By dominated convergence and Jensen's inequality we thus obtain
  \begin{equation*}
    \sup_t \left|\mathbb{P}(Y_n - \theta \le t) - \Phi(t)\right|
    \le \E\left[
      \sup_t \left|F_n(t \mid T_n) - \Phi(t)\right| \right]
    \to 0,
  \end{equation*}
  which gives the result when combined with Slutsky's lemmas.
\end{proof}
%
